Dynamic capabilities and competitive advantage

### 2019-10-26 article4_jbr.rnw
draft intro finished

The presence of a direct path from dynamic capabilities and competetive advantage is
somewhat puzzling. It eiter suggests ommited importrant mediating factors such as
resources, or that dynamic capabilities can act outside "creating, extending and modifying
the firm resources base". A growing consensus on the definition as well as the
theoretical mechanisms underpinning dynamic capabilities suggest that dc work indirectly
towards strategic change by changing firm resources. Hence, the presence of direct effects
seems increasingly odd. Still, seminal contributions to the field of dynamic capabilities have argued for both direct
\cite{Kor2005,Eriksson2014,Schilke2014a} and indirect (ref) effects of dynamic
capabilities on competitive advantage, although increasingly the bulk of the literature
focuses on the latter \citep{Schilke2018}.

 Resource orchestration is defined as "blablbalbla" (helfat)

This indirect link is typically understood as dynamic capabilities modifying, extending or
creating firm resources (helfat) (or lower-level operational capabilities (Winter,
collis)) and thus generating competitive advantage (Eriksson and others).The direct path,
however, is theoretically less explicitly dealt with in the extant literature although
empirical evidence keep suggesting that both direct (ref) and indirect (ref) links exist.

In this paper we explore to what extent a direct link between dynamic capabilities and
competitive advantage work, and under which conditions it is most pronounced. Specifically
the research question is:

{\bf How is a direct link between dynamic capabilities and competitive advantage working
  and under which conditions does it come into play or not?}

We argue that direct link exist as a response to the multilevel notion of strategy itself
and captures strategic change not typically associated with the resource based
view. Furthermore, we show that the prevalence of the direct link is a function of the
nature of underlying resources dynamic capabilities are aimed at "creating, extending and
modifying", specifically if they are valuable and idiosyncratic to the firm or not, and
the environment facing the firm.

Our findings indicate that the direct path is less prevalent in very dynamic environments
where the indirect path becomes far more important. Furthermore, the effect of dynamism is
partly contingent on the nature of the firm resources suggesting a more complex
understanding of how imitable and simple resources can play a role in generating
competetive advantage.

Our contributions are twofold. First, we explore different mediators by explicitly
modeling a causal chain with different types of underlying resources, as well as
under different environments. This is explicitly called for in the literature as "only
four percent of the articles in our sample explicitly examined possible causal mechanisms,
suggesting a significant gap in the literature" (Schilke 2018). This work sheds light on
the contingencies and boundary conditions of the link between dynamic capabilities and the
"holy grail of strategy" (Schilke et al), competetive advantage.

Second, we explore a theoretical explanation for why a direct link between dynamic
capabiliites and competetive advantage may exist. By leaning on empirical evidence
suggesting a multitude of explanations for performance differentials between firms, we
suggest that dyanmic capabilities can act outside the theorized path of "resource
orchestration" through strateic efforts such as industry positioning (porter) or
institutional leverage (Peng).

%%%%%%%%%%%%%%5

%%%%%%%%%%%%%%

%% This warrant the question of \emph{when}
%% direct and indirect links between dynamic capabilities and competitive advantage is most
%% pronounced.

%% Moreover, firm resources or capabiltiies themselves are far from a homogenous set and
%% varies among other things with respect to their uniqueness, inimatbility and value
%% (barney). Important contributions have thus argued for a deeper insighi into these
%% intermediaries and their mechanism calling for a "deeper insight into the variety of mechanisms that
%% underlie the performance effects of capabilities.” (Schilke, 2014a: 199). 
%% as well as \emph{what} different intermediaries exist.

%% This paper investigates the boundary conditions of the VRIO condition and how dynamic
%% capabilities and resources interact to create firm competitive advantage. We ask the
%% following research question:
%% Under which conditions are dynamic capabilities able to yield competietive advantage
%% through resource orchestration and what are the different outcomes of different types of resources.

%% We find thatthe dynamism surrounding hte firm represent an important boundary condition as
%% suggested in parts of the literature. However, DC consinutes to generate competetive
%% advnatage also in high-velocity environments, but doing increasingly so indirectly through
%% resource orchestration. Furthermore, we find that non-vrio resources are more likely to
%% yield competetive power in high velocity environments, and that vrio-resources tend to
%% maintain their CA under any condition.

%% We demonstrate that two paths indeed do exist and that their relative importance varies
%% with the degree of environmental dynamism facing the firm.

%% Under what cionditions are the direct path and indirect path most prevalant? Under which condi

%% The direct

>>>>

Extant literature on dynamic capabilities are increasingly focusing on how firm resources
are changed by dynamic capabilities as a response to changes in the environment
\citep{Schilke2018,Eriksson2014}, thus placing firm resources as a mediator between
dynamic capabilities and competetive advantage \citep{Helfat2007}. Less focus has been
placed on the nature of these underlying firm resources (ref) and how they are capturing a
rather heterogeneous bulk of routines, capabilities, assets and capital. Hence, there is a
need to disentangle the black box of firm resources to provide a more nuanced
understanding of different types of firm resources and how they interact to shape firm
competitive advantage \cite{Schilke2018}. Indeed, this gap in the literature is explicitly
emphasized by \cite{Schilke2018} where they 'see an interesting opportunity for future
work to add greater richness to our understanding of the mechanism of resource-base
change, given this mechanism’s central role in many foundational works (..) and the
diverse ways in which resource changes can potentially come about' (p. 419).

The conventional mechanism explored in the extant literature is related to changes in the
firm`s resource base \citep{Eriksson2014,Schilke2018}. This mechanism is also included in
the dominating definition of dynamic capabilities found in \citep{Helfat2007}: 'A dynamic
capability is the capacity of an organization to purposefully create, extend or modify its
resource base' (p. 4). And although this definition opens up for a variety of components
included in the term 'resource base', a point we discuss later in this paper, most of the
subsequent contributions have focused on changes in operational routines (ref),
capabilities (ref), practices (ref) or tangible assets (ref). This focus, albeit very
important to the development of dynamic capabilities literature, seems to leave out a
considerable part of a firm`s 'resource base', namely those resources that are tied to how
the employees of the firm sees their mission, internalize information and effectuates
change. These traits of the organization is very much related to concepts of
'organizational culture' (ref), 'employee cognition' (ref)' 'organizational genes'
\citep{Nelson1982} and 'sense-making' \citep{Weick1995} to name but a few. However, in
relations with dynamic capabilities one promising avenue for exploring mechanisms beyond
the conventional ones described above, makes a clear distinction between \emph{behavioral}
and \emph{non-behavioral} objects of change \citep{Verona2003}. These non-behavioral
objects of change contain psychological objects such as cognitive frames, motivation and
identity that are important for strategy formation and implementation (Gavetti etc, vince
and gabriel ch 15 of book by verona) as they 'concerns how decisions or actions are
shaped, subverted and transformed by emotions, and (..) determine the "way we do things
here"' (Vince and gabriel). In this respect they work as firm resources in generating
outcomes such as performance and competitive advantage. This notion is perhaps expressed
most clearly by Peter Drucker`s famous quote about how 'culture eats strategy for lunch'
(ref) insofar as . Moreover, these non-behavioral objects of change are themselves far
from static and are found to change over time as responses to external change
\citep{Zollo2016} thus making them a natural part of the mechanisms underpinning how
dynamic capabilities work \citep{Schilke2018,Verona2003}. Still, far less is know about
the non-behavioral objects of change and how they work as a mechanism in the dynamic
capabilities theory.

We aim to contribute to bridging this gap by theorize about how non-behavioral objects of
change work as a mechanism through which dynamic capabilities influence competitive
advantage. Specifically, we ask the following research question:

{\bf What is the role of non-behavioral objects of change as a mechanism through which
  dynamic capabilities can influence competitive advantage?}

To shed light on this important question we theorize on the role of a particular type of
non-behavioral object of change, \emph{organizational cognition}, and how it acts as a
mediator between dynamic capabilities and firm competitive advantage and hence has a
pivotal place in the literature. We build on important work on the concept of
organizational cognition (ref), non-behavioral object of change (ref) and blablabla (ref),
and hypothesize about how organizational cognition acts as a mediator in the same way as
conventional behavioral objects of change (such as routines and capabilities) and that
these two are complementary to a certain degree. Moreover, we also suggest how
organizational cognition could also influence routines and capabilites directly with
further implications for competitive advantage.

This paper makes the following contributions:

The rest of this paper is organized as follows:
